A-maze of Natural Stories: Texts are comprehensible using the Maze task

### 2021-08-18 Papers/Paper/manuscript.Rmd
work on paper

---
title             : "A-maze of Natural Stories: Texts are comprehensible during the Maze task"
shorttitle        : "A-maze of Natural Stories"

@@

abstract: |
  We find support for the localization of reading time effects during the Maze task, as well as extending the range of materials Maze is suitable for.  How long it takes to read a word in a sentence is reflective of how hard it is to identify and integrate the word in the surrounding context. Techniques that slow down the reading process and localize the processing time for each word are useful to understanding the time course of language processing. A-maze is a new method for measuring incremental sentence processing that can localize slowdowns related to syntactic ambiguities. We adapt A-maze for use on longer passages and test it on the Natural Stories corpus. We find that people can comprehend these longer text passages during the Maze task. Moreover, the  task yields useable reaction time data with word predictability effects that are linear in the surprisal of the current word, with little spillover effect from the surprisal of the previous word. This expands the types of effects that can be studied with A-maze, showing it to be a a versatile alternative to eye-tracking and self-paced reading.

@@

r_refs("r-references.bib", append=F)
theme_set(theme_bw())
options(knitr.table.format = "pdf")
```
# Intro
It's remarkable how flexible we are when reading; while we do occasionally stumble when we read something unexpected, we often are able to read slightly unexpected things without a problem. However, these expectations shape how fast we read, even if we don't notice a stumble, unexpected words take longer to process as they force us to rebuild our burgeoning mental model of the sentence. Fortunately for fluent readers and unfortunately for studying language, this process of adjustment is very quick, which makes measures of reading time messy.

Measures of online reading are one way to understand language and how the mind processes language. Many theories of language structure and language processing ground out in predictions about the difficulty of processing words [@bartekSearchOnlineLocality2011]. For instance, the subject v object relative debate includes theories that make fine-grained predictions about which word is how slow -- this needs localized methods to adjudicate it [@grodnerConsequencesSerialNature2005; @staubEyeMovementsProcessing2010; @traxlerProcessingSubjectObject2002]. Other theories such as noisy channel processing also need support from localized word-by-word results [@levyIntegratingSurprisalUncertaininput2011].

Incremental processing methods such as self-paced reading or eye-tracking measure how long someone spends looking at one word before moving on and use that as a proxy for how difficult word was in context. How unexpected a word is correlates with how long it takes to read it and move on [@raynerEffectsFrequencyPredictability2004; @klieglLengthFrequencyPredictability2004]. However, the two major methods of measuring incremental processing both suffer from imprecise localization. In eye-tracking, people read naturally which involves skipping words, jumping ahead and looking back, the dynamics of which make it hard to isolate effects [@raynerEffectsFrequencyPredictability2004; @levyEyeMovementEvidence2009; @frazierMakingCorrectingErrors1982; @raynerEyeMovementsReading1998]. Even when reading order is controlled in self-paced reading, readers may maintain ambiguities about what a word was or means until it is later resolved by context, so it may take multiple words for slowdowns to catch up with them [@macdonaldInteractionLexicalSyntactic1993; @koornneefUseVerbbasedImplicit2006].

One effect of this lack of localization is that reading time for a word is dependent not only on how unexpected a word is, but also how unexpected the previous word is. This is an indication of spillover from the previous word. It's well established for eye-tracking and SPR that RTs are roughly linear in terms of a word's surprisal (negative log probability) [@smithEffectWordPredictability2013; @wilcoxPredictivePowerNeural2020; @goodkindPredictivePowerWord2018; @lukeLimitsLexicalPrediction2016]. Due to spillover effects, on SPR and eye-tracking, there is also a positive linear relationship between the surprisal of a previous word and the RT on the current word -- this is an indication of lack of localization. In addition to suprisal predicting RT, word length and word's overall frequency are also often found to be predictive [@klieglLengthFrequencyPredictability2004].

An alternative method that seems to have superior localization is the Maze task, which adopts an unnatural way of reading to force incremental processing [@freedman85; @forsterMazeTaskMeasuring2009]. In the Maze task, participants see two words at a time, a correct word that continues the sentence, and a distractor which does not. Participants must choose the correct word, and their reaction time (RT) is the dependent measure. If participants make a mistake, the sentence discontinues. Two versions of the maze task exist: lexical or L-maze where the distractors are nonce words and grammatical or G-maze where the distractors are real words of the language that don't fit in the context of the sentence so far. Theoretically, participants must fully integrate each word into the sentence in order to confidently select it. This idea is supported by studies finding strongly localized effects for G-maze which is more sensitive than L-maze [@witzelComparisonsOnlineReading2012a].

The downside of G-maze is that materials are effort-intensive to construct because of the need to select infelicitious words as distractors for each spot of each sentence; this may explain why the Maze task was not widely adopted. @boyceMazeMadeEasy2020 demonstrate a way to automatically generate Maze distractors by using NLP language models to find words that are high-surprisal in the context of the target sentence. The quality of these A-maze distractors is not up to that of hand-generated G-maze distractors, but @boyceMazeMadeEasy2020 found that materials with A-maze distractors had similar results to the hand-generated distractors from @witzelComparisonsOnlineReading2012a. A-maze outperformed L-maze and an SPR control in detecting and localizing expected slowdown effects. @sloggettAmazeAnyOther2020 also found that A-maze and G-maze distractors yielded similar results on a disambiguation paradigm.

A-maze is a potentially powerful addition to the psycholinguists toolkit. However, like the other Maze tasks it has been limited in its application to single-sentence items probing minimal comparisons in constructed sentences. This limits its useful, as some important questions such as comparing human data to language models or studying discourse effects require processing times over multi-sentence passages. Therefore, it's important to expand the Maze task to these types of materials and verify that it finds comparable patterns to other methods.

@@

Using this tweak, we test A-maze on the passages from the Natural Stories corpus. The Natural Stories corpus [@futrellNaturalStoriesCorpus2020] consists of 10 passages each roughly 1000 words long which are designed to read fluently to native speakers. At the same time, the passages contain copious punctuation, quoted speech, many proper nouns, and low frequency grammatical constructions. Taken together, these properties make this a severe test of our process, as these features make it harder for the language model to choose good distractors and require focus from participants. If participants can succeed at the Maze task on this set of materials, we think they are likely to succeed on a wide variety of naturalistic texts. The corpus is accompanied by binary-choice comprehension questions, 6 per story, which we use to assess comprehension. @futrellNaturalStoriesCorpus2020 collected self-paced reading time data on this corpus, which we compare the A-maze results to. Participants were able to read and understand these long passages using A-maze with the error correction paradigm, and their RT profiles showed a linear relationship to the surprisal of the words.

```{r diagram, out.height="25%", fig.width=8, fig.height=3, fig.pos="ht", fig.cap="(ref:diagram-cap)"}
knitr::include_graphics(here("Papers/maze_diagram_cropped.pdf"))
```
One advantage of the Maze task is that it forces incremental processing and automatically excludes inattentive participants by terminating a sentence when a participant makes an error. Thus, only participants who have processed the sentence and are paying attention contribute data at critical regions later in the sentence. However, this means we don't have data after a participant makes a mistake in an item. In traditional G-maze tasks, with hand-crafted distractors and attentive participants, this is a small issue. However, this data loss is much worse with A-maze materials and crowd-sourced participants [@boyceMazeMadeEasy2020]. The high errors are likely from some combination of participants guessing randomly and from auto-generated distractors that in fact fit the sentence; as @boyceMazeMadeEasy2020 noted, some distractors, especially early in the sentence, were problematic and caused considerable data loss.

@@

Participant accuracy reflects both how well participants can navigate the task and what quality the auto-generated distractors are. We calculated the per-word error rate for each participant and graphed it against their average reaction time. (To avoid biasing the average if a participant took a pause before returning to the task, RTs greater than 5 seconds were excluded.) As seen in Figure \@ref(fig:errors)A, one cluster of participants (marked in green) make relatively few errors, with some reaching 99% accuracy. This confirms that the distractors were generally appropriate and shows that some participants maintained focus on the task for the whole story. These careful participants took around 1 second for each word selection, which is much slower than other paradigms. Another cluster of participants (in red) sped through the task, seemingly clicking randomly. This bimodal distribution is likely due to the mix of workers on Mechanical Turk, as we did not use qualification cutoffs.

>>>>

---
title             : "A-maze of Natural Stories: Comprehension and surprisal in the Maze task TODO title"
shorttitle        : "A-maze of Natural Stories"

@@

abstract: |
   TODO needs work
   A-maze is a new method for measuring incremental sentence processing that can localize slowdowns related to syntactic ambiguities. We adapt A-maze for use on longer passages and test it on the Natural Stories corpus. We find that people can comprehend these longer text passages during the Maze task. Moreover, the  task yields useable reaction time data with word predictability effects that are linear in the surprisal of the current word, with little spillover effect from the surprisal of the previous word. This expands the types of effects that can be studied with A-maze, showing it to be a a versatile alternative to eye-tracking and self-paced reading.We find support for the localization of reading time effects during the Maze task, as well as extending the range of materials Maze is suitable for.  How long it takes to read a word in a sentence is reflective of how hard it is to identify and integrate the word in the surrounding context. Techniques that slow down the reading process and localize the processing time for each word are useful to understanding the time course of language processing.

@@

r_refs("r-references.bib", append=F)
theme_set(theme_bw())
options(knitr.table.format = "pdf")
```
# Introduction

Sometimes we stumble in reading, where we expected a sentence to go in one direction and it did not, and then are at a parsing dead-end where we have to read a sentence again. The words that cause stumbles are the outliers on a distribution of words difficulty to read, within fluent reading, there is variation in how easy words are to read and process in their context.  We're able to read many long or surprising words without noticing a problem, but these less expected words do take longer to process as we rebuild our burgeoning mental model of the sentence. Fortunately for fluent readers and unfortunately for studying language, this process of adjustment is very quick, with millisecond scale differences in reading time between words.

We don't have a thorough understanding of how the mind processes language, but we can learn something about what is going on from measuring how long it takes in different circumstances. This relies on a linking hypothesis that processing time reflects something about how much processing is going on, even if we don't know whether this is character recognition, memory retrieval, or parsing. This is enough to compare theories that predict the relative difficulty, and thus relative processing time, of two words in different contexts.

For instance, different theories about what makes object relative clauses harder to understand than subject relative clauses make different predictions about which words are the loci of the overall difficulty and slower reading times associated with object relatives [@grodnerConsequencesSerialNature2005; @staubEyeMovementsProcessing2010; @traxlerProcessingSubjectObject2002]. Measures of reading time can also inform theories about the time course of processing (i.e. which steps are parallel versus serial, @bartekSearchOnlineLocality2011) or the functional form of relationships between word characteristics and processing time [@smithEffectWordPredictability2013].

Some of these theories rely on being able to attribute processing slowdowns to a particular word. Determining that object relatives are overall slower that subject relatives is easy, even an imprecise measure of reading time will determine that the same set of words in a different order took longer to read on a sentence level. However, many language processing theories make specific (and contrasting) theories about which words in a sentence should be harder to process. To adjudicate these theories, we want methods that are *localized*, where it is easy to determine which word is responsible for an observed slow-down in reading time. Ideally, a longer reading time on a word would be an indication of that word's increased difficulty, and not the lingering signal of a prior word's increased difficulty. When the signal isn't localized, advanced analysis techniques may be required to disentangle the slow-downs [@shainDeconvolutionalTimeSeries2018].

## Incremental processing methods

Behavioral methods that measure how long it takes to read each word in a sentence are referred to as incremental processing methods, and their dependent measures are called reading times or reaction times, both abbreviated RT. The two most commonly used methods are eye-tracking and self-paced reading, but both of these suffer from a lack of localization. In eye-tracking, participants read a text on a screen naturally, while their eye-movements are recorded [@raynerEyeMovementsReading1998]. The downside of this natural reading is that people often skip short words and look ahead or look back as they read, the dynamics of which make it hard to isolate effects [@raynerEffectsFrequencyPredictability2004; @levyEyeMovementEvidence2009; @frazierMakingCorrectingErrors1982]. Self-paced reading is a more controlled process where a participant sees one word on screen at a time and presses a button to see the next word instead. Even in this method, readers may maintain ambiguities about what a word was or means until it is later resolved by context, so it may take multiple words for slowdowns to catch up with them [@macdonaldInteractionLexicalSyntactic1993; @koornneefUseVerbbasedImplicit2006]. Both these methods are prone to spillover, as the time to plan motor movements like saccades or button presses is substantial compared to the time it takes to recognize a word, these movements are often initiated before a word is fully processed, so by the time any late processing difficulties occur, the reader may already be several words along.

### the Maze task
An alternative method that is designed to increase localization at the expense of naturalness is the Maze task [@freedman85; @forsterMazeTaskMeasuring2009]. In the Maze task, participants see two words at a time, a correct word that continues the sentence, and a distractor which does not. Participants must choose the correct word, and their time to selection is treated as the reaction time (RT). @forsterMazeTaskMeasuring2009 introduces two versions of the Maze task: lexical L-maze where the distractors are nonce words and grammatical G-maze where the distractors are real words that don't fit with the context of the sentence so far. Theoretically, participants must fully integrate each word into the sentence in order to confidently select it; this may require mentally reparsing previous material in order to allow the integration and selection of a disambiguating word. @forsterMazeTaskMeasuring2009 call this forced incremental processing to distinguish from other incremental processing methods where words can be passively read before later committing to a parse. This idea of strong localization is supported by studies finding strongly localized effects for G-maze which is more sensitive than L-maze [@witzelComparisonsOnlineReading2012a].

The downside of G-maze is that materials are effort-intensive to construct because of the need to select infelicitious words as distractors for each spot of each sentence; this may explain why the Maze task has not been widely adopted. @boyceMazeMadeEasy2020 demonstrate a way to automatically generate Maze distractors by using NLP language models to find words that are high-surprisal in the context of the target sentence, and thus likely to be judged infelicitous by human readers. @boyceMazeMadeEasy2020 found that materials with A-maze distractors had similar results to the hand-generated distractors from @witzelComparisonsOnlineReading2012a. A-maze outperformed L-maze and an SPR control in detecting and localizing expected slowdown effects. @sloggettAmazeAnyOther2020 also found that A-maze and G-maze distractors yielded similar results on a disambiguation paradigm.

Another recent variant of the Maze task is interpolated or I-maze, which uses a mix of real word distractors (generated via the A-maze process) and nonce word distractors [@vaniUsingInterpolatedMaze2021]. The presence of real word distractors encourages close attention to the sentential context, while nonce words can be used as distractors where the word in the sentence is itself ungrammatical or highly unexpected. @vaniUsingInterpolatedMaze2021 used I-maze to compare object- and subject- relative clauses and @wilcoxTargetedAssessmentIncremental2021 used it to compare differences in RT for pairs of grammatical and ungrammatical sentences NLP language models predictions.

## How to compare localization & effects, (old school freq/length), define surprisal and lang models
TODO!!!
One measure of how localized a method is is to look at how much overall effects on each word stay on that word, versus appearing on downstream words. 
There are a few general effects for how properties of words tend to effect their reading time. 
The classic pair are frequency and length -- longer and less common words take longer to process. @klieglLengthFrequencyPredictability2004
Another measure is predictability, or how expected a word is in context, with more predictable words being read faster @raynerEffectsFrequencyPredictability2004. A word can have a high surprisal for lots of reasons -- it could be low frequency, or semantically unexpected, or the start of a low-frequency syntactic construction, or a word that disambiguates prior words to the less common parse. Many targeted effects of interest contribute to how predictable or unpredictable a word is. If effects of predictability are strong and well-localized, it's a good sign. 
Because predictability is made up of all these components, how do we assess how predictable a word is? One way of doing this is used language models that are trained to predict the next word. Predictability is often measured in terms of bits of surprisal, which is the negative log probability of a word (1 bit of surprisal means a word is expected to occur half the time, 2 bits is 1/4 of the time etc). 
RTs for eye-tracking and SPR show a linear relationship between a words surprisal and RT It's well established for eye-tracking and SPR that RTs are roughly linear in terms of a word's surprisal (negative log probability) [@smithEffectWordPredictability2013; @wilcoxPredictivePowerNeural2020; @goodkindPredictivePowerWord2018; @lukeLimitsLexicalPrediction2016].
 Due to spillover effects, on SPR and eye-tracking, there is also a positive linear relationship between the surprisal of a previous word and the RT on the current word -- this is an indication of lack of localization.

## why current experiment
TODO!!!
Maze is great, but there hasn't been a study looking at these coarse effects. We're interested in whether there's spillover ,and how similar it is to SPR. It's only used on targeted/constructed materials of independent sentences.

So we want to do it on some longer natural text and look for these underlying patterns. & see how well participants can read and understand longer texts.

The Natural Stories corpus [@futrellNaturalStoriesCorpus2020] consists of 10 passages each roughly 1000 words long which are designed to read fluently to native speakers. At the same time, the passages contain copious punctuation, quoted speech, many proper nouns, and low frequency grammatical constructions. Taken together, these properties make this a severe test of our process, as these features make it harder for the language model to choose good distractors and require focus from participants. If participants can succeed at the Maze task on this set of materials, we think they are likely to succeed on a wide variety of naturalistic texts. The corpus is accompanied by binary-choice comprehension questions, 6 per story, which we use to assess comprehension.

We tweak the A-maze task to accomodate these longer passages and then have participants read them and also compare with  @futrellNaturalStoriesCorpus2020 collected self-paced reading time data on this corpus, which we compare the A-maze results to.

Using this tweak, we test A-maze on the passages from the Natural Stories corpus.  Participants were able to read and understand these long passages using A-maze with the error correction paradigm, and their RT profiles show a similar pattern to SPR but with less noise and spillover

```{r diagram, out.height="25%", fig.width=8, fig.height=3, fig.pos="ht", fig.cap="(ref:diagram-cap)"}
knitr::include_graphics(here("Papers/maze_diagram_cropped.pdf"))
```
*mention the whole terminate on errors thing*
A-maze is a potentially powerful addition to the psycholinguists toolkit. However, like the other Maze tasks it has been limited in its application to single-sentence items probing minimal comparisons in constructed sentences. This limits its useful, as some important questions such as comparing human data to language models or studying discourse effects require processing times over multi-sentence passages. Therefore, it's important to expand the Maze task to these types of materials and verify that it finds comparable patterns to other methods.

@@

One advantage of the Maze task is that it forces incremental processing and automatically excludes inattentive participants by terminating a sentence when a participant makes an error. Thus, only participants who have processed the sentence and are paying attention contribute data at critical regions later in the sentence. However, this means we don't have data after a participant makes a mistake in an item. In traditional G-maze tasks, with hand-crafted distractors and attentive participants, this is a small issue. However, this data loss is much worse with A-maze materials and crowd-sourced participants [@boyceMazeMadeEasy2020]. The high errors are likely from some combination of participants guessing randomly and from auto-generated distractors that in fact fit the sentence; as @boyceMazeMadeEasy2020 noted, some distractors, especially early in the sentence, were problematic and caused considerable data loss.

@@

```{r}

data_long <- data_filt %>% filter(type!="practice") %>% filter(rt>5000)
data_no_prac <- data_filt %>% filter(type!="practice")
```

Participant accuracy reflects both how well participants can navigate the task and what quality the auto-generated distractors are. We calculated the per-word error rate for each participant and graphed it against their average reaction time. (To avoid biasing the average if a participant took a pause before returning to the task, RTs greater than 5 seconds were excluded; this excluded `r nrow(data_long)` words, or `r nrow(data_long)/nrow(data_no_prac)*100 %>% round(2)`% of trials.) As seen in Figure \@ref(fig:errors)A, one cluster of participants (marked in green) make relatively few errors, with some reaching 99% accuracy. This confirms that the distractors were generally appropriate and shows that some participants maintained focus on the task for the whole story. These careful participants took around 1 second for each word selection, which is much slower than other paradigms. Another cluster of participants (in red) sped through the task, seemingly clicking randomly. This bimodal distribution is likely due to the mix of workers on Mechanical Turk, as we did not use qualification cutoffs.
